---
title: "Student Performance Factors — Business Case"
author: "Wolf & Avery"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)

library(tidyverse)
library(gt)
library(scales)

source("R/load.R")
source("R/preprocess.R")
source("R/missingness.R")
source("R/split.R")
source("R/model.R")
source("R/tune.R")
source("R/eval.R")
source("R/eval_plots.R")
source("R/ensemble.R")
source("R/business.R")
source("R/tables.R")
source("R/theme.R")

ggplot2::theme_set(report_theme())
```

`Modeling.Rmd` compared eight model families (plus a full and a pruned
mean-probability ensemble) on the validation split; the **pruned ensemble** —
`glmnet`, `logistic`, `xgboost`, `lda`, and `rf`, the five highest-AUC models,
mean-pooled after dropping the two weakest (`tree` and `knn`) — had the
highest AUC and F1 there. AUC and F1 don't know about dollars, though, so this
notebook re-fits those five base models plus the ensemble, tunes each one's
*own* decision threshold against the cost assumptions below, and lets **net
benefit** — not AUC — decide which model and threshold actually gets used to
answer the question the model comparison could not: **is running an
intervention program on its predictions worth the money?**

# 1. Cost Assumptions

Everything downstream is driven by two numbers, and both are parameters.

**Edit them here to re-run the entire business case under different assumptions.**

```{r cost-params}
INTERVENTION_COST <- 1000   # spent on every student we flag (TP and FP alike)
NO_ACTION_COST    <- 5000   # what a failure costs when nobody intervenes (FN)
```

The intervention is assumed to work: a flagged student who would have failed
does not. So a true positive converts a `r dollar(NO_ACTION_COST)` failure into
a `r dollar(INTERVENTION_COST)` intervention, netting
`r dollar(NO_ACTION_COST - INTERVENTION_COST)`. A false positive is a wasted
`r dollar(INTERVENTION_COST)`. A false negative eats the full
`r dollar(NO_ACTION_COST)`: the student fails and nothing was spent trying to
stop it. A true negative is free.

```{r payoff-table}
table_payoff(payoff_table(INTERVENTION_COST, NO_ACTION_COST),
             INTERVENTION_COST, NO_ACTION_COST) %>%
  save_table_png("business-payoff")
```

```{r payoff-matrix, fig.width=7, fig.height=6}
plot_payoff_matrix(INTERVENTION_COST, NO_ACTION_COST)
```


# 2. Refit the Winning Model

Same data preparation, same split, same seed as `Modeling.Rmd`: every member
model is fit on **train** only, the threshold is chosen on **validation**
only, and **test** is touched once, at the end.

```{r fit}
CUTOFF <- 63

raw  <- load_("StudentPerformanceFactors.csv")
miss <- count_missing_values(raw)
df   <- impute_categoricals(raw, miss$variable)
df   <- add_target_and_factors(df, cutoff = CUTOFF)
use_linear_contrasts()

num_vars <- select_num_vars(c("Hours_Studied", "Attendance", "Sleep_Hours",
                             "Previous_Scores", "Tutoring_Sessions",
                             "Physical_Activity"))
cat_vars <- select_cat_vars(df)

splits <- split_data(df, c(train = 0.6, val = 0.2, test = 0.2),
                     strata = "Status", seed = 42)

prep_params <- fit_preprocessor(splits$train, num_vars)
train_df <- apply_preprocessor(prep_params, splits$train)
val_df   <- apply_preprocessor(prep_params, splits$val)
test_df  <- apply_preprocessor(prep_params, splits$test)

actual_val  <- as.integer(val_df$Status  == "Fail")
actual_test <- as.integer(test_df$Status == "Fail")
```

The ensemble pools five base learners, each refit here with the hyperparameters
`Modeling.Rmd`'s CV grid search already settled on (`logistic` and `lda` have
none to tune):

| Model | Tuned hyperparameters |
|---|---|
| `glmnet` | `alpha` = 0.1 (`lambda` chosen internally by `cv.glmnet`) |
| `rf` | `mtry` = 2, `ntree` = 600, `nodesize` = 1 |
| `xgboost` | `nrounds` = 300, `max_depth` = 2, `eta` = 0.1, `subsample` = 0.7, `min_child_weight` = 1 |

```{r fit-ensemble}
fit_logistic_m <- fit_logistic(train_df, num_vars, cat_vars)
fit_lda_m      <- fit_lda(train_df, num_vars, cat_vars)
fit_glmnet_m   <- fit_glmnet(train_df, num_vars, cat_vars, alpha = 0.1)
fit_rf_m       <- fit_rf(train_df, num_vars, cat_vars,
                         mtry = 2, ntree = 600, nodesize = 1)
fit_xgboost_m  <- fit_xgboost(train_df, num_vars, cat_vars,
                              nrounds = 300, max_depth = 2, eta = 0.1,
                              subsample = 0.7, min_child_weight = 1)

base_val_probs <- list(
  logistic = predict_probs(fit_logistic_m, val_df, num_vars, cat_vars),
  lda      = predict_probs(fit_lda_m,      val_df, num_vars, cat_vars),
  glmnet   = predict_probs(fit_glmnet_m,   val_df, num_vars, cat_vars),
  rf       = predict_probs(fit_rf_m,       val_df, num_vars, cat_vars),
  xgboost  = predict_probs(fit_xgboost_m,  val_df, num_vars, cat_vars)
)
base_test_probs <- list(
  logistic = predict_probs(fit_logistic_m, test_df, num_vars, cat_vars),
  lda      = predict_probs(fit_lda_m,      test_df, num_vars, cat_vars),
  glmnet   = predict_probs(fit_glmnet_m,   test_df, num_vars, cat_vars),
  rf       = predict_probs(fit_rf_m,       test_df, num_vars, cat_vars),
  xgboost  = predict_probs(fit_xgboost_m,  test_df, num_vars, cat_vars)
)

pruned_names <- prune_by_auc(base_val_probs, actual_val, top_n = 5)
pruned_names

ensemble_val_probs  <- ensemble_probs(base_val_probs[pruned_names])
ensemble_test_probs <- ensemble_probs(base_test_probs[pruned_names])

all_val_probs  <- c(base_val_probs,  list(ensemble = ensemble_val_probs))
all_test_probs <- c(base_test_probs, list(ensemble = ensemble_test_probs))
```

# 3. Choose the Model and Decision Rule on Validation

`Modeling.Rmd` picked a winner by AUC/F1 alone, using a single shared
threshold-tuning metric across models. Here every one of the six candidates
(the five base learners plus the ensemble) gets its *own* shot at four
candidate decision rules, each producing one threshold on the validation set:

| Rule | Threshold from | Optimizes |
|---|---|---|
| `argmax` | fixed at 0.5 | nothing — the textbook "predict the more likely class" default |
| `f1` | tuned on validation | harmonic mean of precision and recall |
| `youdens_j` | tuned on validation | sensitivity + specificity − 1 |
| `cost_ratio` | `r dollar(INTERVENTION_COST)` / `r dollar(NO_ACTION_COST)` = `r round(INTERVENTION_COST / NO_ACTION_COST, 3)` | expected cost, from the assumptions alone — flag a student whenever `p x NO_ACTION_COST > INTERVENTION_COST` |

The first three are statistical rules that ignore the cost asymmetry; the
fourth is derived from it. For each model, only its best-net-benefit rule is
kept; the models are then ranked against each other at that best rule.
Whichever **model + rule** combination produces the highest **validation** net
benefit is the one carried to test.

```{r rules}
model_rules <- best_rule_per_model(all_val_probs, actual_val,
                                   INTERVENTION_COST, NO_ACTION_COST)
table_model_rule_comparison(model_rules) %>%
  save_table_png("business-model-rule-comparison")
```

```{r best-pick}
best_model     <- model_rules$model[1]
best_rule      <- model_rules$rule[1]
best_threshold <- model_rules$threshold[1]
c(model = best_model, rule = best_rule, threshold = best_threshold)
```

```{r rules-curve, fig.width=8, fig.height=5}
best_model_rules <- threshold_rules(all_val_probs[[best_model]], actual_val,
                                    INTERVENTION_COST, NO_ACTION_COST)
plot_net_benefit_curve(all_val_probs[[best_model]], actual_val, best_model_rules,
                       INTERVENTION_COST, NO_ACTION_COST)
```

```{r rule-pick-note, echo=FALSE, results='asis'}
tied <- model_rules$model[model_rules$net_benefit == max(model_rules$net_benefit)]
tie_note <- if (length(tied) > 1) {
  sprintf(" (%s tie exactly; `%s` is taken as the first)",
          paste(sprintf("`%s`", tied), collapse = " and "), best_model)
} else ""

cat(sprintf(
"The **%s** model wins on validation with its **%s** rule at a threshold of **%.3f**%s, worth %s against the %s the do-nothing baseline would cost.\n",
best_model, best_rule, best_threshold, tie_note,
dollar(max(model_rules$net_benefit)),
dollar(model_rules$cost_no_action[1])))
```

# 4. Apply to Test

The model and the threshold are both now fixed. This is the only time the test
set is scored.

```{r test-eval}
test_probs <- all_test_probs[[best_model]]
nb_test <- net_benefit(test_probs, actual_test, best_threshold,
                       INTERVENTION_COST, NO_ACTION_COST)
nb_test
```

```{r test-cm, fig.width=7, fig.height=6}
plot_cost_confusion_matrix(nb_test, INTERVENTION_COST, NO_ACTION_COST,
                           title = sprintf("Test Set in Dollars — %s, %s rule",
                                          best_model, best_rule))
```

```{r test-scorecard}
table_business_case(nb_test, label = "Test set") %>%
  save_table_png("business-scorecard")
```

```{r test-waterfall, fig.width=8, fig.height=5}
plot_business_case(nb_test)
```

# 5. Bottom Line

```{r bottom-line, echo=FALSE, results='asis'}
roi <- nb_test$net_benefit / nb_test$investment
cat(sprintf(
"- Doing nothing costs **%s** across the %d students in the test set (%d of them fail at %s each).\n- The program flags **%d students** (%.1f%% of the cohort) for a total investment of **%s**.\n- It still misses **%d failures**, a residual loss of **%s**.\n- Total cost with the model is **%s** against **%s** without it: a net benefit of **%s**, or %s per student and %s returned per dollar spent.\n",
dollar(nb_test$cost_no_action), nb_test$n, nb_test$tp + nb_test$fn, dollar(NO_ACTION_COST),
nb_test$n_flagged, 100 * nb_test$n_flagged / nb_test$n, dollar(nb_test$investment),
nb_test$fn, dollar(nb_test$loss),
dollar(nb_test$total_cost), dollar(nb_test$cost_no_action),
dollar(nb_test$net_benefit), dollar(nb_test$net_benefit_per_student),
dollar(roi)))
```

Two caveats worth stating plainly:

- **The intervention is assumed to be perfectly effective.** Every true
  positive is credited with averting a failure. If it only works, say, 60% of
  the time, the value of a true positive falls from
  `r dollar(NO_ACTION_COST - INTERVENTION_COST)` to
  `r dollar(0.6 * NO_ACTION_COST - INTERVENTION_COST)` and the case weakens
  proportionally. Both costs are parameters at the top of this notebook, so
  that scenario is a one-line edit.
- **The data are close to separation** (see `EDA.Rmd`, section 12a), which
  pushes the linear models (`logistic`, `glmnet`, `lda`) toward overconfident
  probabilities near 0 and 1; if the ensemble is the winner, averaging them
  with `rf` and `xgboost` tempers this somewhat, but not entirely. That mostly
  affects rules that read the probability scale literally, such as
  `cost_ratio`; the net-benefit curve above is flat enough across the middle
  of the range that the exact threshold matters less than the decision to run
  the program at all.
