---
title: "Student Performance Factors — EDA & Failure Modeling"
author: "Wolf & Avery"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)

library(tidyverse)
library(GGally)
library(gt)
library(scales)
library(patchwork)
library(broom)
library(car)

source("R/load.R")
source("R/preprocess.R")
source("R/missingness.R")
source("R/eda_plots.R")
source("R/tables.R")
source("R/model.R")
source("R/diagnostics.R")

ggplot2::theme_set(report_theme())
```

# 1. Load & Inspect

```{r load}
raw <- load_("StudentPerformanceFactors.csv")
```

# 2. Pass/Fail Threshold Selection

Before anything else, we pick the Pass/Fail cutoff. The plot below shows the
distribution of `Exam_Score` with a cumulative "% would pass" line; the
dashed line marks the grade needed to hit a target pass rate (`target_prop`).
The table on the right lists each candidate threshold (`thresholds`) alongside
its cumulative pass/fail rates, bolding whichever grade comes closest to
`target_prop`.

```{r threshold-tuning, fig.width=10, fig.height=5}
plot_pass_fail_threshold_tuning(
  raw, 
  thresholds = c(60, 61, 62, 63, 64, 65, 66, 67, 68, 69, 70), 
  target_prop = 0.9
  )
```

`CUTOFF` drives every downstream Pass/Fail split in this notebook — change it
here to re-run the EDA at a different threshold.

```{r cutoff-param}
CUTOFF <- 63
```

# 3. Missingness

```{r missing}
miss <- count_missing_values(raw)
tabulate_missing_values(miss)
```

# 4. Imputation

Target (`Exam_Score`) is complete, so no rows are dropped. The three affected
categoricals are mode-imputed.

```{r impute}

df <- impute_categoricals(raw, miss$variable)
```

## 4a. Outlier Check

`Exam_Score` is nominally a 0-100 scale; the boxplot below flags any value
above that range (e.g. a `101`, likely a data entry error or bonus points),
which is then dropped.

```{r exam-outliers}
plot_exam_score_outliers(df)
```

```{r exam-outliers-filter}
nrow(df)
df <- filter_exam_outliers(df)
nrow(df)
```

# 5. Pass/Fail Target (cutoff = `r CUTOFF`) + Factor Setup

Ordinal predictors get explicit level ordering (Low → High) so effects read
monotonically; the rest become plain factors.

```{r target}
df <- add_target_and_factors(df, cutoff = CUTOFF)
table_class_balance(df, cutoff = CUTOFF) %>%
  save_table_png("eda-class-balance")
```

```{r target-plot}
plot_target_balance(df)
```

# 6. Univariate Distributions

Sections 6-8 look at every candidate predictor; the ones that survive into the
model are chosen in section 9.

## 6a. Numeric

```{r dist-num}
num_vars_all <- c("Hours_Studied","Attendance","Sleep_Hours","Previous_Scores",
                  "Tutoring_Sessions","Physical_Activity")
plot_num_distributions(df, num_vars_all)
```

```{r dist-num-box}
plot_num_boxplots(df, num_vars_all)
```

## 6b. Categorical

```{r dist-cat, fig.height=9}
cat_vars_all <- df %>% select(where(is.factor), -Status) %>% names()
plot_cat_distributions(df, cat_vars_all)
```

# 7. Associations with Failing

## 7a. Numeric predictors vs. Status

```{r box-num}
plot_num_by_status(df, num_vars_all)
```

```{r ttests}
smd <- ttest_summary(df, num_vars_all)
table_ttest(smd) %>%
  save_table_png("eda-ttest")
```

## 7b. Categorical predictors vs. Status

```{r cat-assoc}
fail_rates <- fail_rate_summary(df, cat_vars_all)
table_fail_rates(fail_rates) %>%
  save_table_png("eda-fail-rates")
```

```{r cat-plot, fig.height=9}
plot_fail_rates_cat(df, cat_vars_all)
```

```{r cat-plot-stacked, fig.height=9}
plot_status_by_cat_stacked(df, cat_vars_all)
```

```{r chisq}
chi_res <- chisq_summary(df, cat_vars_all)
table_chisq(chi_res) %>%
  save_table_png("eda-chisq")
```

## 7c. Interaction Screening

`Hours_Studied` split by `Parental_Involvement` hints at a possible
interaction — `Low` involvement behaves differently from `Medium`/`High`.
The dot chart below generalizes this check across every categorical
predictor, ranking each level by its standardized `Hours_Studied` gap
between Pass and Fail; wide spread within a variable flags it as a
candidate interaction term.

```{r interaction-density}
plot_hours_by_status_interaction(df)
```

```{r interaction-gap, fig.height=9}
gap_results <- interaction_gap_summary(df, "Hours_Studied", cat_vars_all)
plot_interaction_gap(gap_results)
```

# 8. Correlation Structure

```{r corr, fig.height=7}
plot_correlation(df, num_vars_all)
```

```{r corr-pairs, fig.width=10, fig.height=10}
plot_num_pairs(df, num_vars_all)
```

# 9. Feature Selection

An initial kitchen-sink fit on all 19 predictors made three things clear, and
the screens above agree with all three:

- **`Sleep_Hours`** (z = -1.38), **`School_Type`** (z = 0.88) and **`Gender`**
  (z = 0.52) contributed nothing. Their fail rates are flat across levels
  (section 7b) and `Sleep_Hours` is the one numeric predictor whose Pass/Fail
  boxplots overlap almost completely (section 7a). All three are dropped.
- **Every quadratic (`.Q`) contrast** was non-significant (the largest was
  `Access_to_Resources.Q`, borderline at p = 0.05; the rest had |z| < 2). The
  fail-rate bars in section 7b step monotonically across
  `Low → Medium → High`, so a single linear (`.L`) trend per ordinal is the
  right parameterization — this drops 8 df with essentially no loss of fit.
- **`Motivation_Level`** was being read as an unordered factor, so it got dummy
  coding while the other seven ordinals got polynomial contrasts. It is now an
  ordered factor (`R/load.R`) and enters as a single `.L` term like its peers.

`use_linear_contrasts()` implements the second point by making a linear-only
polynomial contrast the default coding for ordered factors;
`select_num_vars()`/`select_cat_vars()` implement the first by removing
`PRUNED_VARS` (see `R/preprocess.R`).

```{r feature-selection}
use_linear_contrasts()

num_vars <- select_num_vars(num_vars_all)
cat_vars <- select_cat_vars(df)

PRUNED_VARS
c(num_vars, cat_vars)
```

# 10. Logistic Regression

Model **P(Fail)**; positive coefficients raise failure risk. `Exam_Score` is
excluded (it defines the target). Each ordinal predictor contributes one `.L`
coefficient reading Low→High, so a negative `.L` means risk falls as the
predictor rises.

```{r glm-fit}
fit <- fit_logistic(df, num_vars, cat_vars)
summary(fit)
```

## 10a. Coefficients Table

```{r coef-table}
table_coef(fit) %>%
  save_table_png("eda-coef")
```

## 10b. Log-Odds Forest Plot

```{r logodds-plot, fig.height=7}
plot_logodds(fit)
```

```{r fit-quality}
table_fit_quality(fit) %>%
  save_table_png("eda-fit-quality")
```

# 11. Interaction Screening

Section 7c raised one visual hint of an interaction (`Hours_Studied` behaving
differently under `Low` parental involvement). This is the formal version of
that check: refit the additive model above with one two-way interaction added
at a time and compare the two with a likelihood-ratio test. With
`r choose(length(c(num_vars, cat_vars)), 2)` candidate terms, raw p-values
alone would throw off several false positives, so p-values are
Benjamini-Hochberg adjusted.

```{r interaction-screen}
int_screen <- screen_interactions(df, num_vars, cat_vars)
table_interaction_lrt(int_screen) %>%
  save_table_png("eda-interaction-lrt")
```

```{r interaction-screen-summary, echo=FALSE, results='asis'}
n_raw <- sum(int_screen$p_value < 0.05, na.rm = TRUE)
n_adj <- sum(int_screen$p_adj   < 0.05, na.rm = TRUE)
cat(sprintf(
"Nothing survives: %d of %d terms clear the raw p < 0.05 bar, against the ~%.0f expected by chance alone, and **%d** clear the BH-adjusted bar. The section 7c signal was sampling noise in a thin slice of the data, so the model stays additive.\n",
n_raw, nrow(int_screen), 0.05 * nrow(int_screen), n_adj))
```

# 12. Assumption Checks

## 12a. Logistic Regression

Logistic regression needs independent observations, a logit that is linear in
each numeric predictor, no severe multicollinearity, and no handful of points
driving the fit. Independence holds by construction (one row per student, no
repeated measures or clustering variable in the data), so the remaining three
are checked below.

**Linearity of the logit.** Binning each numeric predictor and plotting the
observed log-odds of failing per bin should trace a straight line.

```{r logit-linearity, fig.height=6}
plot_empirical_logit(df, num_vars)
```

**Residuals.** Individual logistic residuals are uninformative, so residuals
are averaged within bins of the fitted probability; each point should sit
inside its own ±2 SE bar with no pattern across bins.

```{r binned-resid}
plot_binned_residuals(fit)
```

The bins collapse onto 0 and 1 rather than spreading across the x-axis, which
is the separation problem showing up again: the model assigns almost every
student a fitted probability of essentially 0 or 1, so there are only a handful
of distinct values left to bin. Within those clumps the mean residuals are
small and unpatterned, but the plot has little power to detect misspecification
here.

**Multicollinearity.** Generalized VIFs, converted to the ordinary VIF scale so
the usual "> 5" rule applies.

```{r vif}
table_vif(vif_summary(fit)) %>%
  save_table_png("eda-vif")
```

`Hours_Studied` and `Attendance` come back above 20, which would normally be
alarming — but the raw predictors are essentially uncorrelated (section 8: every
off-diagonal |r| < 0.03). GVIF is computed from the IRLS-weighted design, and
those weights are `p(1-p)`, which collapse to ~0 for almost every row once the
model is nearly separated. So this is a symptom of the separation diagnosed
below, not genuine collinearity between the two predictors.

**Influential observations.** Cook's distance against the conventional 4/n
screening cut-off.

```{r influence-plot}
plot_influence(fit)
```

```{r influence-table}
table_influence(influence_summary(fit), n_obs = nrow(df)) %>%
  save_table_png("eda-influence")
```

**Separation.** Failing is the minority class and the fitted coefficients are
large, which is the signature of near-complete separation — the condition under
which the MLE drifts toward infinity and standard errors stop being
trustworthy. This is the single most important caveat on the coefficients
above.

```{r separation}
table_separation(separation_summary(fit)) %>%
  save_table_png("eda-separation")
```

## 12b. LDA / QDA

LDA and QDA both assume the predictors are multivariate normal *within* each
class; LDA additionally assumes both classes share one covariance matrix, which
is exactly the assumption QDA relaxes. Only the numeric predictors can be
assessed — the 11 retained categoricals cannot be normal by construction, which
is already a structural violation for both methods.

**Within-class normality.**

```{r lda-density, fig.height=6}
plot_within_class_density(df, num_vars)
```

```{r lda-qq, fig.width=11, fig.height=5}
plot_within_class_qq(df, num_vars)
```

```{r lda-shapiro}
table_normality(normality_summary(df, num_vars)) %>%
  save_table_png("eda-normality")
```

**Equal covariance (LDA-specific).** Levene's test per predictor, then Box's M
on the full covariance matrices, then a look at *where* the two structures
differ.

```{r lda-levene}
table_levene(levene_summary(df, num_vars)) %>%
  save_table_png("eda-levene")
```

```{r lda-boxm}
table_boxm(boxm_test(df, num_vars)) %>%
  save_table_png("eda-boxm")
```

```{r lda-cor, fig.width=10, fig.height=5}
plot_class_correlations(df, num_vars)
```

Expect QDA to handle this data better than LDA, and treat both as weaker than
the tree/ensemble models in `Modeling.Rmd`, which carry no distributional
assumptions at all.

# 13. Takeaways

```{r takeaways, echo=FALSE, results='asis'}
top_num <- smd$variable[1]
top_cat <- chi_res$variable[1]
cat(sprintf(
"- Class balance: %.1f%% fail (failing is the minority class here).\n- Strongest numeric signal: **%s**.\n- Strongest categorical signal: **%s**.\n- Missingness was light (three categoricals, ~1%%) and mode-imputed.\n- `Exam_Score` must stay out of any predictive model.\n- Feature selection took the kitchen-sink fit (19 predictors, 27 coefficients) down to %d predictors and %d coefficients: `Sleep_Hours`, `School_Type` and `Gender` dropped outright, and each ordinal reduced to a single linear trend.\n- No interaction survived screening, so the model stays additive.\n- Because attendance and study hours dominate, an intervention model should probably weight those over the socioeconomic factors.\n- Caveat: the fit sits close to complete separation, so coefficient *directions* are reliable but their *magnitudes* are not; LDA's equal-covariance assumption is also violated.\n",
100 * mean(df$Status == "Fail"), top_num, top_cat,
length(c(num_vars, cat_vars)), length(coef(fit))))
```
