---
title: "Student Performance Factors — EDA & Failure Modeling"
author: "Wolf & Avery"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)

library(tidyverse)
library(GGally)
library(gt)
library(scales)
library(patchwork)
library(broom)

source("R/load.R")
source("R/preprocess.R")
source("R/missingness.R")
source("R/eda_plots.R")
source("R/tables.R")
source("R/model.R")

ggplot2::theme_set(report_theme())
```

# 1. Load & Inspect

```{r load}
raw <- load_("StudentPerformanceFactors.csv")
```

# 2. Pass/Fail Threshold Selection

Before anything else, we pick the Pass/Fail cutoff. The plot below shows the
distribution of `Exam_Score` with a cumulative "% would pass" line; the
dashed line marks the grade needed to hit a target pass rate (`target_prop`).
The table on the right lists each candidate threshold (`thresholds`) alongside
its cumulative pass/fail rates, bolding whichever grade comes closest to
`target_prop`.

```{r threshold-tuning, fig.width=10, fig.height=5}
plot_pass_fail_threshold_tuning(
  raw, 
  thresholds = c(60, 61, 62, 63, 64, 65, 66, 67, 68, 69, 70), 
  target_prop = 0.9
  )
```

`CUTOFF` drives every downstream Pass/Fail split in this notebook — change it
here to re-run the EDA at a different threshold.

```{r cutoff-param}
CUTOFF <- 63
```

# 3. Missingness

```{r missing}
miss <- count_missing_values(raw)
tabulate_missing_values(miss)
```

# 4. Imputation

Target (`Exam_Score`) is complete, so no rows are dropped. The three affected
categoricals are mode-imputed.

```{r impute}

df <- impute_categoricals(raw, miss$variable)
```

## 4a. Outlier Check

`Exam_Score` is nominally a 0-100 scale; the boxplot below flags any value
above that range (e.g. a `101`, likely a data entry error or bonus points),
which is then dropped.

```{r exam-outliers}
plot_exam_score_outliers(df)
```

```{r exam-outliers-filter}
nrow(df)
df <- filter_exam_outliers(df)
nrow(df)
```

# 5. Pass/Fail Target (cutoff = `r CUTOFF`) + Factor Setup

Ordinal predictors get explicit level ordering (Low → High) so effects read
monotonically; the rest become plain factors.

```{r target}
df <- add_target_and_factors(df, cutoff = CUTOFF)
table_class_balance(df, cutoff = CUTOFF)
```

```{r target-plot}
plot_target_balance(df)
```

# 6. Univariate Distributions

## 6a. Numeric

```{r dist-num}
num_vars <- c("Hours_Studied","Attendance","Sleep_Hours","Previous_Scores",
              "Tutoring_Sessions","Physical_Activity")
plot_num_distributions(df, num_vars)
```

```{r dist-num-box}
plot_num_boxplots(df, num_vars)
```

## 6b. Categorical

```{r dist-cat, fig.height=9}
cat_vars <- df %>% select(where(is.factor), -Status) %>% names()
plot_cat_distributions(df, cat_vars)
```

# 7. Associations with Failing

## 7a. Numeric predictors vs. Status

```{r box-num}
plot_num_by_status(df, num_vars)
```

```{r ttests}
smd <- ttest_summary(df, num_vars)
table_ttest(smd)
```

## 7b. Categorical predictors vs. Status

```{r cat-assoc}
fail_rates <- fail_rate_summary(df, cat_vars)
table_fail_rates(fail_rates)
```

```{r cat-plot, fig.height=9}
plot_fail_rates_cat(df, cat_vars)
```

```{r cat-plot-stacked, fig.height=9}
plot_status_by_cat_stacked(df, cat_vars)
```

```{r chisq}
chi_res <- chisq_summary(df, cat_vars)
table_chisq(chi_res)
```

## 7c. Interaction Screening

`Hours_Studied` split by `Parental_Involvement` hints at a possible
interaction — `Low` involvement behaves differently from `Medium`/`High`.
The dot chart below generalizes this check across every categorical
predictor, ranking each level by its standardized `Hours_Studied` gap
between Pass and Fail; wide spread within a variable flags it as a
candidate interaction term.

```{r interaction-density}
plot_hours_by_status_interaction(df)
```

```{r interaction-gap, fig.height=9}
gap_results <- interaction_gap_summary(df, "Hours_Studied", cat_vars)
plot_interaction_gap(gap_results)
```

# 8. Correlation Structure

```{r corr, fig.height=7}
plot_correlation(df, num_vars)
```

```{r corr-pairs, fig.width=10, fig.height=10}
plot_num_pairs(df, num_vars)
```

**NOTE:** We considered applying PCA to the numeric predictors, but the correlation
analysis above shows negligible pairwise correlation among all six numeric
variables (all |r| < 0.03). Since PCA's value comes from compressing
correlated, redundant information into fewer components, and our predictors
show minimal redundancy, we determined PCA would not meaningfully reduce
dimensionality or add value here, and chose not to pursue it further.

# 9. Logistic Regression

Model **P(Fail)**; positive coefficients raise failure risk. `Exam_Score` is
excluded (it defines the target). Ordinal factors use polynomial contrasts, so
`.L` (linear) terms capture the Low→High trend.

```{r glm-fit}
fit <- fit_logistic(df, num_vars, cat_vars)
summary(fit)
```

## 9a. Coefficients Table

```{r coef-table}
table_coef(fit)
```

## 9b. Log-Odds Forest Plot

```{r logodds-plot, fig.height=7}
plot_logodds(fit)
```

```{r fit-quality}
table_fit_quality(fit)
```

# 10. Takeaways

```{r takeaways, echo=FALSE, results='asis'}
top_num <- smd$variable[1]
top_cat <- chi_res$variable[1]
cat(sprintf(
"- Class balance: %.1f%% fail (failing is the minority class here).\n- Strongest numeric signal: **%s**.\n- Strongest categorical signal: **%s**.\n- Missingness was light (three categoricals, ~1%%) and mode-imputed.\n- `Exam_Score` must stay out of any predictive model.\n- Because attendance and study hours dominate, an intervention model should probably weight those over the socioeconomic factors.\n",
100 * mean(df$Status == "Fail"), top_num, top_cat))
```
