---
title: "Student Performance Factors — EDA & Failure Modeling"
author: "Wolf & Avery"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)

library(tidyverse)
library(GGally)
library(gt)
library(scales)
library(patchwork)
library(broom)

source("R/load.R")
source("R/preprocess.R")
source("R/missingness.R")
source("R/eda_plots.R")
source("R/tables.R")
source("R/model.R")

ggplot2::theme_set(report_theme())
```

# 1. Load & Inspect

```{r load}
raw <- load_("StudentPerformanceFactors.csv")
```

# 2. Missingness

```{r missing}
miss <- count_missing_values(raw)
tabulate_missing_values(miss)
```

# 3. Imputation

Target (`Exam_Score`) is complete, so no rows are dropped. The three affected
categoricals are mode-imputed.

```{r impute}

df <- impute_categoricals(raw, miss$variable)
```

# 4. Pass/Fail Target (cutoff = 60) + Factor Setup

Ordinal predictors get explicit level ordering (Low → High) so effects read
monotonically; the rest become plain factors.

```{r target}
df <- add_target_and_factors(df)
table_class_balance(df)
```

```{r target-plot}
plot_target_balance(df)
```

# 5. Univariate Distributions

```{r dist-num}
num_vars <- c("Hours_Studied","Attendance","Sleep_Hours","Previous_Scores",
              "Tutoring_Sessions","Physical_Activity")
plot_num_distributions(df, num_vars)
```

# 6. Associations with Failing

## 6a. Numeric predictors vs. Status

```{r box-num}
plot_num_by_status(df, num_vars)
```

```{r ttests}
smd <- ttest_summary(df, num_vars)
table_ttest(smd)
```

## 6b. Categorical predictors vs. Status

```{r cat-assoc}
cat_vars <- df %>% select(where(is.factor), -Status) %>% names()

fail_rates <- fail_rate_summary(df, cat_vars)
table_fail_rates(fail_rates)
```

```{r cat-plot, fig.height=9}
plot_fail_rates_cat(df, cat_vars)
```

```{r chisq}
chi_res <- chisq_summary(df, cat_vars)
table_chisq(chi_res)
```

# 7. Correlation Structure

```{r corr, fig.height=7}
plot_correlation(df, num_vars)
```

# 8. Logistic Regression

Model **P(Fail)**; positive coefficients raise failure risk. `Exam_Score` is
excluded (it defines the target). Ordinal factors use polynomial contrasts, so
`.L` (linear) terms capture the Low→High trend.

```{r glm-fit}
fit <- fit_logistic(df, num_vars, cat_vars)
summary(fit)
```

## 8a. Coefficients Table

```{r coef-table}
table_coef(fit)
```

## 8b. Log-Odds Forest Plot

```{r logodds-plot, fig.height=7}
plot_logodds(fit)
```

```{r fit-quality}
table_fit_quality(fit)
```

# 9. Takeaways

```{r takeaways, echo=FALSE, results='asis'}
top_num <- smd$variable[1]
top_cat <- chi_res$variable[1]
cat(sprintf(
"- Class balance: %.1f%% fail (failing is the majority class here).\n- Strongest numeric signal: **%s**.\n- Strongest categorical signal: **%s**.\n- Missingness was light (three categoricals, ~1%%) and mode-imputed.\n- `Exam_Score` must stay out of any predictive model.\n- Because attendance and study hours dominate, an intervention model should probably weight those over the socioeconomic factors.\n",
100 * mean(df$Status == "Fail"), top_num, top_cat))
```
