---
title: "Student Performance Factors — Interpretation"
author: "Avery & Wolf"
date: "2026-08-09"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 1. Setup
```{r}
library(tidyverse)
library(ggplot2)
library(car)

source("R/load.R")
source("R/preprocess.R")
source("R/split.R")
source("R/missingness.R")
source("R/model.R")

CUTOFF <- 63

raw <- load_("StudentPerformanceFactors.csv")
miss <- count_missing_values(raw)
df <- impute_categoricals(raw, miss$variable)
df <- add_target_and_factors(df, cutoff = CUTOFF)
use_linear_contrasts()

num_vars <- select_num_vars(c("Hours_Studied", "Attendance", "Sleep_Hours",
                               "Previous_Scores", "Tutoring_Sessions",
                               "Physical_Activity"))
cat_vars <- select_cat_vars(df)
```

# 2. Train / Validation / Test Split
```{r}
splits <- split_data(df, c(train = 0.6, val = 0.2, test = 0.2),
                      strata = "Status", seed = 42)
```

# Additive Model Equation
$$
\log\left(\frac{P(\text{Fail})}{1-P(\text{Fail})}\right) = \beta_0 
+ \beta_1(\text{Hours\_Studied}) 
+ \beta_2(\text{Attendance}) 
+ \beta_3(\text{Previous\_Scores}) 
+ \beta_4(\text{Tutoring\_Sessions}) 
+ \beta_5(\text{Physical\_Activity}) 
+ \beta_6(\text{Parental\_Involvement}) 
+ \beta_7(\text{Access\_to\_Resources}) 
+ \beta_8(\text{Extracurricular\_Activities}) 
+ \beta_9(\text{Motivation\_Level}) 
+ \beta_{10}(\text{Internet\_Access}) 
+ \beta_{11}(\text{Family\_Income}) 
+ \beta_{12}(\text{Teacher\_Quality}) 
+ \beta_{13}(\text{Peer\_Influence}) 
+ \beta_{14}(\text{Learning\_Disabilities}) 
+ \beta_{15}(\text{Parental\_Education\_Level}) 
+ \beta_{16}(\text{Distance\_from\_Home})
$$
\
Following feature selection, `Sleep_Hours`, `School_Type`, and `Gender`
were dropped from the additive model, and all ordinal predictors (Parental_Involvement, Access_to_Resources, Motivation_Level, Family_Income, Teacher_Quality, Peer_Influence, Parental_Education_Level, Distance_from_Home) now enter via a single linear (`.L`) contrast rather than separate linear and
quadratic terms, since the quadratic components were uniformly
non-significant. The remaining categorical predictors
(Extracurricular_Activities, Internet_Access, Learning_Disabilities)
are binary and contribute one dummy-coded coefficient each.\

# 3. Additive Logistic Model (Objective 1)

**Note:** this model is fit on the training split specifically (rather than 
the full dataset used in EDA.Rmd's exploratory fit), so the coefficients 
interpreted here describe the same model whose performance is reported 
in Modeling.Rmd's comparison table.

```{r}
fit <- fit_logistic(splits$train, num_vars, cat_vars)
summary(fit)
```

# 4. Extract coefficients and confidence intervals
```{r}
# Full coefficient table with CIs
coefs <- summary(fit)$coefficients
cis <- confint.default(fit)  

results <- data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  std_error = coefs[, "Std. Error"],
  p_value = coefs[, "Pr(>|z|)"],
  ci_lower = cis[, 1],
  ci_upper = cis[, 2]
)

# Pull one continous and one categorical
results %>% filter(term %in% c("Hours_Studied", "Extracurricular_ActivitiesYes"))
```

# 5. Convert log odds to odds ratio
```{r}
results %>%
  filter(term %in% c("Hours_Studied", "Extracurricular_ActivitiesYes")) %>%
  mutate(
    odds_ratio = exp(estimate),
    or_ci_lower = exp(ci_lower),
    or_ci_upper = exp(ci_upper)
  )
```

# 6. Effects Plot - Hours_Studied
```{r}
# Build a "typical" student profile, varying only Hours_Studied
newdata <- splits$train %>%
  summarise(across(all_of(num_vars), median),
            across(all_of(cat_vars), ~ names(sort(table(.), decreasing = TRUE))[1])) %>%
  slice(rep(1, 50)) %>%
  mutate(Hours_Studied = seq(min(splits$train$Hours_Studied), 
                              max(splits$train$Hours_Studied), length.out = 50))

newdata$predicted_prob_fail <- predict(fit, newdata = newdata, type = "response")

ggplot(newdata, aes(x = Hours_Studied, y = predicted_prob_fail)) +
  geom_line(color = "steelblue", linewidth = 1) +
  theme_minimal() +
  labs(title = "Predicted Probability of Failing vs. Hours Studied",
       subtitle = "Other predictors held at their median/most common values",
       y = "Predicted P(Fail)", x = "Hours Studied")
```

# 7. Effects Plot - Extracurricular_Activities (Adjusted)
```{r}
# Note: evaluated at Hours_Studied = 6 (near the risk transition zone) rather than the training median, since at typical study levels the model predicts near-zero failure risk for all students regardless of Extracurricular_Activities

newdata_extra2 <- splits$train %>%
  summarise(across(all_of(num_vars), median),
            across(all_of(cat_vars), ~ names(sort(table(.), decreasing = TRUE))[1])) %>%
  slice(rep(1, 2)) %>%
  mutate(Extracurricular_Activities = c("No", "Yes"),
         Hours_Studied = 6)  # near the risky zone, not the overall median

pred2 <- predict(fit, newdata = newdata_extra2, type = "link", se.fit = TRUE)

newdata_extra2 <- newdata_extra2 %>%
  mutate(
    predicted_prob_fail = plogis(pred2$fit),
    ci_lower = plogis(pred2$fit - 1.96 * pred2$se.fit),
    ci_upper = plogis(pred2$fit + 1.96 * pred2$se.fit)
  )

ggplot(newdata_extra2, aes(x = Extracurricular_Activities, y = predicted_prob_fail)) +
  geom_point(size = 4, color = "steelblue") +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, color = "steelblue") +
  theme_minimal() +
  labs(title = "Predicted Probability of Failing by Extracurricular Activities",
       subtitle = "Evaluated at Hours_Studied = 6 (near risk transition zone)",
       y = "Predicted P(Fail)", x = "Extracurricular Activities")

newdata_extra2 %>% dplyr::select(Extracurricular_Activities, predicted_prob_fail, ci_lower, ci_upper)
```

# 8. Interpretation (Coefficient / Odds Ratio / Confidence Intervals)

**Hours_Studied:** Holding all other predictors constant, each additional hour studied is associated with a decrease in the log-odds of failing by 1.77 (SE = 0.18, p < 0.001). Converting the log odds to an odds ratio, each additional hour studied is associated with roughly an 83% reduction in the odds of failing (OR = 0.17, 95% CI: 0.12–0.24).\
\
**Extracurricular_Activities:** Holding all other predictors constant, students who participate in extracurricular activities have odds of failing about 89% lower than those who don't (OR = 0.11, 95% CI: 0.05–0.26). This is equivalent to a decrease in the log-odds of failing by 2.17 (SE = 0.42, p < 0.001).\


# 9. Practical vs Statistical Significance

**Hours_Studied:** While the statistical significance of this effect is clear, its practical magnitude raises a cause for concern. The effects plot shows an unusually sharp transition in predicted failure probability and near-certain failure below roughly 5 hours studied, collapsing to near-certain passing at above roughly 10 hours studied, with the entire shift occurring in a narrow window. Notably, this steepness persisted even after removing weak predictors and simplifying the model's contrasts (which improved AIC from 270.6 to 262.6). This suggests that the sharpness reflects a genuine pattern of separation near the Fail threshold in this data, rather than an artifact of unnecessary model complexity.\ 
A relationship this deterministic is unlikely to translate in the real world. It more likely reflects some residual separation in the model, a byproduct of how rare Fail cases are in this dataset even after adjusting our threshold. We consider the direction of this effect (more hours studied associated with lower risk of failing) to be reliable and well-supported, but we recommend treating the exact magnitude of the effect with caution rather than as a generalizable estimate.\
\
**Extracurricular_Activities:** To assess practical significance of Extracurricular_Activities, we evaluated predicted failure probability for a student in a borderline study-time range (6 hours studied — near the model's steepest risk transition), since at typical study levels the predicted risk of failing is already near zero for all students regardless of extracurricular participation.\ 
At this borderline study-time range, predicted probability of failing drops from 79.6% (95% CI: 54.8%–92.6%) for non-participants of extracurricular activities to 30.7% (95% CI: 12.9%–56.9%) for participants — roughly a 49-percentage-point reduction. This is a substantial and practically meaningful effect, particularly for students already at elevated risk based on study time. The relatively wide confidence intervals reflect genuine uncertainty in this estimate, so this would be a limitation worth acknowledging rather than obscuring.\

# 10. Broader Predictor Importance

A note on method first: since our coefficients mix unstandardized continuous 
variables (Hours_Studied) with binary/ordinal dummy variables, raw coefficient 
size isn't directly comparable across predictors. The best available comparison 
from a standard glm() output is the z-value (coefficient ÷ SE), which reflects 
how confidently each effect is estimated relative to its own uncertainty — a 
reasonable proxy for relative importance here.\

```{r}
results_full <- data.frame(
  term = rownames(coefs),
  z = abs(coefs[, "z value"])
) %>% arrange(desc(z))

results_full
```

Across the 16 retained predictors, Attendance (z = 9.86) and Hours_Studied (z = 9.79) stand out as the two strongest signals in the model. This is consistent with our univariate EDA, where these same two variables showed the largest mean differences between Pass and Fail students (Attendance: −14.4 points, p < 0.0001; Hours_Studied: −6.7 hours, p < 0.0001). `Previous_Scores`, `Access_to_Resources`, and `Tutoring_Sessions` round out the top five, which also coincide with their strong standalone associations with failing identified in the EDA stage.\
\
Among the categorical predictors, `Access_to_Resources` and `Parental_Involvement` carry the most weight — again matching the two strongest chi-square associations found in EDA (χ² = 49.2 and 77.1, respectively). This consistency between our univariate screening and the final multivariate model gives us more confidence that these aren't spurious or interaction-driven effects, but genuinely robust predictors of failure risk.\
\
Nearly all coefficients point in the protective direction (higher levels associated with lower failure risk). However, there are two notable exceptions worth flagging:\
\
`Learning_Disabilities` (positive coefficient — presence of a learning disability is associated with higher failure risk) and `Distance_from_Home` (positive .L — farther distance from home is associated with higher risk, potentially tied to reduced attendance or engagement). Both align with intuitive, real-world expectations rather than contradicting them.\
`Extracurricular_Activities` — one of the two worked examples above — has the smallest |z| of the entire 16 (z = 5.16), and it is the weakest signal in the model. It remains statistically significant and, as our effects plot showed, still practically meaningful in the higher-risk range of `Hours_Studied`. However, it's worth noting that it ranks last among the retained predictors, not first, despite its featured role in Section 8.\

# 11. LDA / QDA Assumptions
```{r}
# Visual Check (Distributions of Numeric Variables)
splits$train %>%
  select(all_of(num_vars), Status) %>%
  pivot_longer(-Status, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = Status)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Numeric Predictor Distributions by Status",
       subtitle = "Checking approximate within-class normality for LDA/QDA")
```

```{r}
levene_results <- map_dfr(num_vars, function(v) {
  test <- leveneTest(splits$train[[v]] ~ splits$train$Status)
  tibble(
    variable = v,
    f_value = test$`F value`[1],
    p_value = test$`Pr(>F)`[1]
  )
})

levene_results
```

# Write Up:

LDA and QDA both assume that predictor variables are approximately normally distributed within each class. LDA additionally assumes the covariance structure of the predictors is equal across classes (QDA is more relaxed on this second assumption).\
\
**Normality:** Of our 19 total predictors, 13 are categorical and cannot satisfy a normality assumption by definition. Among the 6 numeric predictors, visual inspection of within-class density plots shows mixed results: `Physical_Activity`, `Sleep_Hours`, and `Tutoring_Sessions` are discrete count variables with multimodal, spiky distributions that are clearly non-normal. `Attendance` and `Previous_Scores` show skewed, non-bell-shaped distributions that differ in shape between classes. `Hours_Studied` is the best-behaved of the six, showing an approximately unimodal, roughly symmetric distribution in both classes, though it does display a mild right skew.\
\
**Equal covariance (LDA-specific):** Levene's test, run separately on each numeric predictor, shows the equal-variance assumption is violated for several key variables. `Attendance` shows an extreme difference in variance between the Fail and Pass classes (F = 250.0, p < 0.001), with `Tutoring_Sessions` (F = 17.0, p < 0.001) and `Hours_Studied` (F = 12.8, p < 0.001) also showing significant, though more modest, violations. `Sleep_Hours`, `Physical_Activity`, and `Previous_Scores` do not show statistically significant differences in variance between classes. Overall, the equal-covariance assumption required by LDA appears violated for roughly half of the numeric predictors, most severely for `Attendance` which is one of the model's strongest predictors. These results are suggesting that LDA's core assumption is not well-supported with this data.\
\
**Conclusion:** Both the normality and equal-covariance assumptions appear substantially violated in this dataset, primarily driven by the large number of categorical predictors, and further supported by the skewed/discrete shapes and unequal spread among the numeric predictors. Given QDA doesn't assume equal-covariance, we might reasonably expect it to outperform LDA here. However, both models' core normality assumption remains questionable given the predominance of categorical predictors, so results from either should be interpreted cautiously relative to models that don't carry these distributional assumptions.\
\

# Scope and Logistical Considerations

**Assumptions about future data:** This model assumes that future students are drawn from a population similar to the one used for training — same school district, similar demographic and socioeconomic composition, and a similar underlying relationship between these predictors and academic performance. It also assumes the definition of "failing" stays consistent at a fixed exam score cutoff of 63. If the assessment itself changes, or if the school's definition of at-risk shifts, the model would need to be re-calibrated rather than applied as-is. Because Fail cases made up only 8.8% of our training data, the model has relatively few real examples of what is considered "underperforming", so predictions should be treated with more caution for students whose profiles fall outside the range commonly seen in the training data.\
\
**Model Instability from Class Balance:** As discussed in the Practical vs. Statistical Significance sections above, several coefficients in this model — particularly `Hours_Studied` — show clear signs of separation, likely a consequence of how rare Fail cases are even after threshold-tuning. This means the model can be overconfident in narrow bands of the predictor space (e.g., predicting near-0% or near-100% failure probability for students just below or above a study-time threshold). In practice, predictions near these extreme probability regions should be treated as directional signals rather than precise risk scores.\
\
**Threshold sensitivity and the cost of errors:** This model feeds into a real intervention decision with asymmetric costs. Regarding the business case guiding this project, a missed at-risk student (false negative) costs approximately 50,000 dollars, while an unnecessary intervention (false positive) costs about 10,000 dollars — a 5:1 cost ratio. This means the decision threshold should be deliberately tuned toward higher Sensitivity, even at the expense of more false positives, rather than optimized for a "balanced" metric like overall accuracy. Any deployment of this model should revisit the threshold if these cost assumptions change (e.g., if intervention costs rise, or if the number of available intervention slots is constrained).\
\
**Practical deployment concerns:** Since the school appears to be identifying a fixed proportion of students (the "bottom 10%") for intervention, this model would need periodic re-validation as new cohorts of students are assessed. Student populations, course rigor, or grading practices could all shift over time in ways that change what an exam score of 63 actually represents. We'd also recommend monitoring the model's predicted Fail rate against the actual observed Fail rate each term, to catch any drift early rather than assuming the original training-time relationships continue to hold indefinitely.
