---
title: "Student Performance Factors — Interpretation"
author: "Avery & Wolf"
date: "2026-08-09"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 1. Setup
```{r}
library(tidyverse)
library(ggplot2)
library(car)

source("R/load.R")
source("R/preprocess.R")
source("R/split.R")
source("R/missingness.R")
source("R/model.R")

CUTOFF <- 63

raw <- load_("StudentPerformanceFactors.csv")
miss <- count_missing_values(raw)
df <- impute_categoricals(raw, miss$variable)
df <- add_target_and_factors(df, cutoff = CUTOFF)

num_vars <- c("Hours_Studied", "Attendance", "Sleep_Hours", "Previous_Scores",
              "Tutoring_Sessions", "Physical_Activity")
cat_vars <- df %>% select(where(is.factor), -Status) %>% names()
```

# 2. Train / Validation / Test Split
```{r}
splits <- split_data(df, c(train = 0.6, val = 0.2, test = 0.2),
                      strata = "Status", seed = 42)
```

# Additive Model Equation
$$
\log\left(\frac{P(\text{Fail})}{1-P(\text{Fail})}\right) = \beta_0 
+ \beta_1(\text{Hours\_Studied}) 
+ \beta_2(\text{Attendance}) 
+ \beta_3(\text{Sleep\_Hours}) 
+ \beta_4(\text{Previous\_Scores}) 
+ \beta_5(\text{Tutoring\_Sessions}) 
+ \beta_6(\text{Physical\_Activity}) 
+ \beta_7(\text{Parental\_Involvement}) 
+ \beta_8(\text{Access\_to\_Resources}) 
+ \beta_9(\text{Extracurricular\_Activities}) 
+ \beta_{10}(\text{Motivation\_Level}) 
+ \beta_{11}(\text{Internet\_Access}) 
+ \beta_{12}(\text{Family\_Income}) 
+ \beta_{13}(\text{Teacher\_Quality}) 
+ \beta_{14}(\text{School\_Type}) 
+ \beta_{15}(\text{Peer\_Influence}) 
+ \beta_{16}(\text{Learning\_Disabilities}) 
+ \beta_{17}(\text{Parental\_Education\_Level}) 
+ \beta_{18}(\text{Distance\_from\_Home}) 
+ \beta_{19}(\text{Gender})
$$
\
For the ordinal predictors (Parental_Involvement, Access_to_Resources,
Motivation_Level, Family_Income, Teacher_Quality, Peer_Influence,
Parental_Education_Level, Distance_from_Home), each contributes two
terms via polynomial contrasts — a linear (`.L`) and quadratic (`.Q`)
component — rather than a single coefficient. The remaining categorical
predictors contribute one coefficient per non-reference level (standard
dummy coding).\
\

# 3. Additive Logistic Model (Objective 1)
```{r}
fit <- fit_logistic(splits$train, num_vars, cat_vars)
summary(fit)
```

# 4. Extract coefficients and confidence intervals
```{r}
# Full coefficient table with CIs
coefs <- summary(fit)$coefficients
cis <- confint.default(fit)  

results <- data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  std_error = coefs[, "Std. Error"],
  p_value = coefs[, "Pr(>|z|)"],
  ci_lower = cis[, 1],
  ci_upper = cis[, 2]
)

# Pull one continous and one categorical
results %>% filter(term %in% c("Hours_Studied", "Extracurricular_ActivitiesYes"))
```

# 5. Convert log odds to odds ratio
```{r}
results %>%
  filter(term %in% c("Hours_Studied", "Extracurricular_ActivitiesYes")) %>%
  mutate(
    odds_ratio = exp(estimate),
    or_ci_lower = exp(ci_lower),
    or_ci_upper = exp(ci_upper)
  )
```

# 6. Effects Plot - Hours_Studied
```{r}
# Build a "typical" student profile, varying only Hours_Studied
newdata <- splits$train %>%
  summarise(across(all_of(num_vars), median),
            across(all_of(cat_vars), ~ names(sort(table(.), decreasing = TRUE))[1])) %>%
  slice(rep(1, 50)) %>%
  mutate(Hours_Studied = seq(min(splits$train$Hours_Studied), 
                              max(splits$train$Hours_Studied), length.out = 50))

newdata$predicted_prob_fail <- predict(fit, newdata = newdata, type = "response")

ggplot(newdata, aes(x = Hours_Studied, y = predicted_prob_fail)) +
  geom_line(color = "steelblue", linewidth = 1) +
  theme_minimal() +
  labs(title = "Predicted Probability of Failing vs. Hours Studied",
       subtitle = "Other predictors held at their median/most common values",
       y = "Predicted P(Fail)", x = "Hours Studied")
```

# 7. Effects Plot - Extracurricular_Activities (Adjusted)
```{r}
# Note: evaluated at Hours_Studied = 6 (near the risk transition zone) rather than the training median, since at typical study levels the model predicts near-zero failure risk for all students regardless of Extracurricular_Activities

newdata_extra2 <- splits$train %>%
  summarise(across(all_of(num_vars), median),
            across(all_of(cat_vars), ~ names(sort(table(.), decreasing = TRUE))[1])) %>%
  slice(rep(1, 2)) %>%
  mutate(Extracurricular_Activities = c("No", "Yes"),
         Hours_Studied = 6)  # near the risky zone, not the overall median

pred2 <- predict(fit, newdata = newdata_extra2, type = "link", se.fit = TRUE)

newdata_extra2 <- newdata_extra2 %>%
  mutate(
    predicted_prob_fail = plogis(pred2$fit),
    ci_lower = plogis(pred2$fit - 1.96 * pred2$se.fit),
    ci_upper = plogis(pred2$fit + 1.96 * pred2$se.fit)
  )

ggplot(newdata_extra2, aes(x = Extracurricular_Activities, y = predicted_prob_fail)) +
  geom_point(size = 4, color = "steelblue") +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, color = "steelblue") +
  theme_minimal() +
  labs(title = "Predicted Probability of Failing by Extracurricular Activities",
       subtitle = "Evaluated at Hours_Studied = 6 (near risk transition zone)",
       y = "Predicted P(Fail)", x = "Extracurricular Activities")

newdata_extra2 %>% dplyr::select(Extracurricular_Activities, predicted_prob_fail, ci_lower, ci_upper)
```

# 8. Interpretation (Coefficient / Odds Ratio / Confidence Intervals)
\
**Hours_Studied:**\
\
Holding all other predictors constant, each additional hour studied is associated with a decrease in the log-odds of failing by 1.82, a highly statistically significant effect (p < 0.001). Converting to an odds ratio, this corresponds to the odds of failing multiplying by a factor of 0.16 (95% CI: 0.11–0.24) for each additional hour studied. In other words, as study time increases, it results in a large, statistically significant reduction in the odds of failing.\
\
**Extracurricular_Activities:**\
\
Holding all other predictors constant, students who participate in extracurricular activities have an odds of failing approximately 0.10 times (95% CI: 0.04–0.24) those of students who do not participate. This is equivalent to a decrease in the log-odds of failing by a factor of 2.29, and it is a highly statistically significant effect (p < 0.001).\
\
\
# 9. Practical vs Statistical Significance
\
**Hours_Studied:**\
\
While the statistical significance of this effect is clear, its practical magnitude raises a cause for concern. The effects plot shows an unusually sharp transition in predicted failure probability — near-certain failure below roughly 5 hours studied, collapsing to near-certain passing at above roughly 10 hours studied, with the entire shift occurring in a narrow window.\ 
A relationship this deterministic is unlikely to translate in the real world. It more likely reflects some residual separation in the model, a byproduct of how rare Fail cases are in this dataset even after adjusting our threshold. We consider the direction of this effect (more hours studied associated with lower risk of failing) to be reliable and well-supported, but recommend treating the exact magnitude of the effect with caution rather than as a precise, generalizable estimate.\
\
**Extracurricular_Activities:**\
\
To assess practical significance of Extracurricular_Activities, we evaluated predicted failure probability for a student in a borderline study-time range (6 hours studied — near the model's steepest risk transition), since at typical study levels the predicted risk of failing is already near zero for all students regardless of extracurricular participation.\ 
At this borderline point, predicted probability of failing drops from 84.3% (95% CI: 53.3%–96.2%) for non-participants to 35.1% (95% CI: 11.6%–69.0%) for participants — a roughly 49-percentage-point reduction. This is a substantial, practically meaningful effect, particularly for students already at elevated risk based on study time. The relatively wide confidence intervals reflect genuine uncertainty in this estimate, since few observations fall within this narrow, high-risk band of the data. This would be a limitation worth acknowledging rather than obscuring.\
\
\
# 10. LDA / QDA Assumptions
```{r}
# Visual Check (Distributions of Numeric Variables)
splits$train %>%
  select(all_of(num_vars), Status) %>%
  pivot_longer(-Status, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = Status)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Numeric Predictor Distributions by Status",
       subtitle = "Checking approximate within-class normality for LDA/QDA")
```

```{r}
levene_results <- map_dfr(num_vars, function(v) {
  test <- leveneTest(splits$train[[v]] ~ splits$train$Status)
  tibble(
    variable = v,
    f_value = test$`F value`[1],
    p_value = test$`Pr(>F)`[1]
  )
})

levene_results
```
\
# Write Up:
\
LDA and QDA both assume that predictor variables are approximately normally distributed within each class. LDA additionally assumes the covariance structure of the predictors is equal across classes (QDA is more relaxed on this second assumption).\
\
**Normality:** Of our 19 total predictors, 13 are categorical and cannot satisfy a normality assumption by definition. Among the 6 numeric predictors, visual inspection of within-class density plots shows mixed results: `Physical_Activity`, `Sleep_Hours`, and `Tutoring_Sessions` are discrete count variables with multimodal, spiky distributions that are clearly non-normal. `Attendance` and `Previous_Scores` show skewed, non-bell-shaped distributions that differ in shape between classes. `Hours_Studied` is the best-behaved of the six, showing an approximately unimodal, roughly symmetric distribution in both classes, though it does display a mild right skew.\
\
**Equal covariance (LDA-specific):** Levene's test, run separately on each numeric predictor, shows the equal-variance assumption is violated for several key variables. `Attendance` shows an extreme difference in variance between the Fail and Pass classes (F = 250.0, p < 0.001), with `Tutoring_Sessions` (F = 17.0, p < 0.001) and `Hours_Studied` (F = 12.8, p < 0.001) also showing significant, though more modest, violations. `Sleep_Hours`, `Physical_Activity`, and `Previous_Scores` do not show statistically significant differences in variance between classes. Overall, the equal-covariance assumption required by LDA appears violated for roughly half of the numeric predictors, most severely for `Attendance` which is one of the model's strongest predictors. These results are suggesting that LDA's core assumption is not well-supported with this data.\
\
**Conclusion:** Both the normality and equal-covariance assumptions appear substantially violated in this dataset, primarily driven by the large number of categorical predictors, and further supported by the skewed/discrete shapes and unequal spread among the numeric predictors. Given QDA doesn't assume equal-covariance, we might reasonably expect it to outperform LDA here. However, both models' core normality assumption remains questionable given the predominance of categorical predictors, so results from either should be interpreted cautiously relative to models that don't carry these distributional assumptions.
