---
title: "Student Performance Factors — Modeling"
author: "Wolf & Avery"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 6)

library(tidyverse)
library(gt)
library(pROC)

source("R/load.R")
source("R/preprocess.R")
source("R/missingness.R")
source("R/split.R")
source("R/model.R")
source("R/tune.R")
source("R/eval.R")
source("R/eval_plots.R")
source("R/tables.R")
source("R/theme.R")

ggplot2::theme_set(report_theme())
```

# 1. Load & Prepare Data

```{r load}
raw <- load_("StudentPerformanceFactors.csv")
miss <- count_missing_values(raw)
df <- impute_categoricals(raw, miss$variable)
df <- add_target_and_factors(df)

num_vars <- c("Hours_Studied", "Attendance", "Sleep_Hours", "Previous_Scores",
              "Tutoring_Sessions", "Physical_Activity")
cat_vars <- df %>% select(where(is.factor), -Status) %>% names()
```

# 2. Train / Validation / Test Split

Stratified on `Status` so every split preserves the Pass/Fail ratio.

```{r split}
splits <- split_data(df, c(train = 0.6, val = 0.2, test = 0.2),
                      strata = "Status", seed = 42)
sapply(splits, nrow)
```

# 3. Scale / Impute Numeric Predictors

Fit on train only; apply the same train-derived mean/sd to validation and test.

```{r preprocess}
prep_params <- fit_preprocessor(splits$train, num_vars)
train_df <- apply_preprocessor(prep_params, splits$train)
val_df   <- apply_preprocessor(prep_params, splits$val)
test_df  <- apply_preprocessor(prep_params, splits$test)
```

# 4. Model Specs & Hyperparameter Search Spaces

Each model provides a `fit_fn(train_df, ...hyperparams)`; models with a
`search_space` are tuned via `grid_search()`, the rest are fit directly.
All models predict through the shared `predict_fn` (dispatches on model
class via `predict_probs()`).

```{r model-specs}
predict_fn <- function(model, d) predict_probs(model, d, num_vars, cat_vars)

model_specs <- list(
  logistic = list(
    fit_fn = function(train_df) fit_logistic(train_df, num_vars, cat_vars),
    search_space = NULL
  ),
  logistic_complex = list(
    fit_fn = function(train_df) {
      fit_logistic_complex(train_df, num_vars, cat_vars,
        c("Hours_Studied:Parental_Involvement", "Hours_Studied:Motivation_Level"))
    },
    search_space = NULL
  ),
  lda = list(
    fit_fn = function(train_df) fit_lda(train_df, num_vars, cat_vars),
    search_space = NULL
  ),
  qda = list(
    fit_fn = function(train_df) fit_qda(train_df, num_vars, cat_vars),
    search_space = NULL
  ),
  glmnet = list(
    fit_fn = function(train_df, alpha) fit_glmnet(train_df, num_vars, cat_vars, alpha = alpha),
    search_space = list(alpha = c(0, 0.25, 0.5, 0.75, 1))
  ),
  tree = list(
    fit_fn = function(train_df, cp) {
      fit_tree(train_df, num_vars, cat_vars, control = rpart::rpart.control(cp = cp))
    },
    search_space = list(cp = c(0.001, 0.005, 0.01, 0.05))
  ),
  rf = list(
    fit_fn = function(train_df, mtry, ntree) {
      fit_rf(train_df, num_vars, cat_vars, mtry = mtry, ntree = ntree)
    },
    search_space = list(mtry = c(2, 4, 6), ntree = c(200, 400))
  ),
  xgboost = list(
    fit_fn = function(train_df, nrounds, max_depth) {
      fit_xgboost(train_df, num_vars, cat_vars, nrounds = nrounds, max_depth = max_depth)
    },
    search_space = list(nrounds = c(50, 100), max_depth = c(3, 6))
  )
)
```

# 5. Fit Each Model on Train, Evaluate on Validation

KNN is handled separately below (`class::knn` has no persisted model
object, so it can't share the generic `fit_fn`/`predict_fn` interface).

```{r fit-and-tune}
fitted_models <- list()
val_probs <- list()
best_hp <- list()

for (name in names(model_specs)) {
  spec <- model_specs[[name]]

  if (is.null(spec$search_space)) {
    model <- spec$fit_fn(train_df)
  } else {
    gs <- grid_search(train_df, val_df, spec$fit_fn, predict_fn, spec$search_space)
    best_row <- gs[gs$best, , drop = FALSE][1, ]
    hp <- as.list(best_row[names(spec$search_space)])
    best_hp[[name]] <- hp
    model <- do.call(spec$fit_fn, c(list(train_df), hp))
  }

  fitted_models[[name]] <- model
  val_probs[[name]] <- predict_fn(model, val_df)
}

best_hp
```

## KNN

```{r knn}
knn_probs <- function(train_df, target_df, num_vars, cat_vars, k) {
  build_x <- function(d) {
    model_df <- d %>% dplyr::select(dplyr::all_of(num_vars), dplyr::all_of(cat_vars))
    stats::model.matrix(~ . - 1, data = model_df)
  }
  train_x  <- build_x(train_df)
  target_x <- build_x(target_df)
  train_cl <- factor(dplyr::if_else(train_df$Status == "Fail", 1L, 0L))

  pred <- class::knn(train_x, target_x, cl = train_cl, k = k, prob = TRUE)
  p <- attr(pred, "prob")
  ifelse(pred == "1", p, 1 - p)
}

actual_val <- as.integer(val_df$Status == "Fail")

k_grid <- c(3, 5, 7, 9, 11, 15)
k_auc <- sapply(k_grid, function(k) {
  probs <- knn_probs(train_df, val_df, num_vars, cat_vars, k)
  as.numeric(pROC::auc(actual_val, probs, quiet = TRUE))
})
best_k <- k_grid[which.max(k_auc)]
best_hp$knn <- list(k = best_k)

val_probs$knn <- knn_probs(train_df, val_df, num_vars, cat_vars, best_k)
```

## Validation Performance (all models, default 0.5 threshold)

```{r val-comparison}
val_comparison <- compare_models(val_probs, actual_val)
table_model_comparison(val_comparison)
```

# 6. Threshold Tuning (Validation Set)

Per-model F1-optimal threshold, chosen on validation only.

```{r threshold-tuning}
thresholds <- purrr::map_dbl(val_probs, ~ optimal_threshold(.x, actual_val, metric = "youdens_j"))
thresholds
```

# 7. Evaluate All Models on Test (Optimal Thresholds)

```{r test-probs}
test_probs <- purrr::imap(fitted_models, function(model, name) predict_fn(model, test_df))
test_probs$knn <- knn_probs(train_df, test_df, num_vars, cat_vars, best_k)

actual_test <- as.integer(test_df$Status == "Fail")
test_comparison <- compare_models(test_probs, actual_test, thresholds = thresholds)
table_model_comparison(test_comparison)
```

# 8. Confusion Matrix — Best Model

```{r best-model-cm}
best_name <- test_comparison$model[which.max(test_comparison$f1)]
best_threshold <- thresholds[[best_name]]

best_eval <- evaluate_predictions(test_probs[[best_name]], actual_test, threshold = best_threshold)
plot_confusion_matrix(best_eval$confusion_matrix,
                       title = paste0("Confusion Matrix — ", best_name,
                                       " (threshold = ", round(best_threshold, 2), ")"))
```

# 9. ROC Curves — All Models

```{r roc-all, fig.height=7}
plot_roc(test_probs, actual_test)
```
