---
title: "Student Performance Factors — Modeling"
author: "Wolf & Avery"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 6)

library(tidyverse)
library(gt)
library(pROC)

source("R/load.R")
source("R/preprocess.R")
source("R/missingness.R")
source("R/split.R")
source("R/model.R")
source("R/tune.R")
source("R/eval.R")
source("R/eval_plots.R")
source("R/importance.R")
source("R/ensemble.R")
source("R/tables.R")
source("R/theme.R")

ggplot2::theme_set(report_theme())
```

# 1. Pass/Fail Threshold

`CUTOFF` sets the Pass/Fail split used for every model below; see `EDA.Rmd`
for the threshold-tuning plot used to choose it.

```{r cutoff-param}
CUTOFF <- 63
```

# 2. Load & Prepare Data

```{r load}
raw <- load_("StudentPerformanceFactors.csv")
miss <- count_missing_values(raw)
df <- impute_categoricals(raw, miss$variable)
df <- add_target_and_factors(df, cutoff = CUTOFF)

num_vars <- c("Hours_Studied", "Attendance", "Sleep_Hours", "Previous_Scores",
              "Tutoring_Sessions", "Physical_Activity")
cat_vars <- df %>% select(where(is.factor), -Status) %>% names()
```

# 3. Train / Validation / Test Split

Stratified on `Status` so every split preserves the Pass/Fail ratio.

```{r split}
splits <- split_data(df, c(train = 0.6, val = 0.2, test = 0.2),
                      strata = "Status", seed = 42)
sapply(splits, nrow)
```

# 4. Scale / Impute Numeric Predictors

Fit on train only; apply the same train-derived mean/sd to validation and test.

```{r preprocess}
prep_params <- fit_preprocessor(splits$train, num_vars)
train_df <- apply_preprocessor(prep_params, splits$train)
val_df   <- apply_preprocessor(prep_params, splits$val)
test_df  <- apply_preprocessor(prep_params, splits$test)
```

# 5. Define the Search Space

Each model provides a `fit_fn(train_df, ...hyperparams)`; models with a
`search_space` are tuned via stratified 5-fold CV on train (`cv_grid_search()`),
the rest are fit directly. All models predict through the shared `predict_fn`
(dispatches on model class via `predict_probs()`).

```{r model-specs}
predict_fn <- function(model, d) predict_probs(model, d, num_vars, cat_vars)

model_specs <- list(
  logistic = list(
    fit_fn = function(train_df) fit_logistic(train_df, num_vars, cat_vars),
    search_space = NULL
  ),
  logistic_complex = list(
    fit_fn = function(train_df) {
      fit_logistic_complex(train_df, num_vars, cat_vars,
        c("Hours_Studied:Parental_Involvement", "Hours_Studied:Motivation_Level"))
    },
    search_space = NULL
  ),
  lda = list(
    fit_fn = function(train_df) fit_lda(train_df, num_vars, cat_vars),
    search_space = NULL
  ),
  qda = list(
    fit_fn = function(train_df) fit_qda(train_df, num_vars, cat_vars),
    search_space = NULL
  ),
  glmnet = list(
    fit_fn = function(train_df, alpha) fit_glmnet(train_df, num_vars, cat_vars, alpha = alpha),
    search_space = list(alpha = c(0, 0.25, 0.5, 0.75, 1))
  ),
  tree = list(
    fit_fn = function(train_df, cp) {
      fit_tree(train_df, num_vars, cat_vars, control = rpart::rpart.control(cp = cp))
    },
    search_space = list(cp = c(0.001, 0.005, 0.01, 0.05))
  ),
  rf = list(
    fit_fn = function(train_df, mtry, ntree) {
      fit_rf(train_df, num_vars, cat_vars, mtry = mtry, ntree = ntree)
    },
    search_space = list(mtry = c(2, 4, 6), ntree = c(200, 400))
  ),
  xgboost = list(
    fit_fn = function(train_df, nrounds, max_depth) {
      fit_xgboost(train_df, num_vars, cat_vars, nrounds = nrounds, max_depth = max_depth)
    },
    search_space = list(nrounds = c(50, 100), max_depth = c(3, 6))
  )
)
```

# 6. Hyperparameter Tuning (CV Grid Search)

For each model with a `search_space`, hyperparameters are selected by
stratified 5-fold CV on `splits$train` (raw/unscaled — `cv_grid_search()`
refits the numeric preprocessor on each fold's training rows only, to
avoid leakage into the held-out fold). The winning hyperparameters are then
used to fit one final model on the full (globally scaled) `train_df`.
Validation predictions from that final model are kept for threshold tuning
below. KNN is handled separately (`class::knn` has no persisted model
object, so it can't share the generic `fit_fn`/`predict_fn` interface).

```{r fit-and-tune}
fitted_models <- list()
val_probs <- list()
best_hp <- list()
cv_grids <- list()

for (name in names(model_specs)) {
  spec <- model_specs[[name]]

  if (is.null(spec$search_space)) {
    model <- spec$fit_fn(train_df)
  } else {
    cv <- cv_grid_search(splits$train, num_vars, cat_vars, spec$fit_fn,
                          predict_fn, spec$search_space, k = 5, seed = 42)
    cv_grids[[name]] <- cv
    best_row <- cv[cv$best, , drop = FALSE][1, ]
    hp <- as.list(best_row[names(spec$search_space)])
    best_hp[[name]] <- hp
    model <- do.call(spec$fit_fn, c(list(train_df), hp))
  }

  fitted_models[[name]] <- model
  val_probs[[name]] <- predict_fn(model, val_df)
}

best_hp
```

## KNN

`k` is selected by the same stratified 5-fold CV procedure (`cv_select_knn_k()`),
then the final model predicts on `val_df` using the full `train_df`.

```{r knn}
actual_val <- as.integer(val_df$Status == "Fail")

k_grid <- c(3, 5, 7, 9, 11, 15)
knn_cv <- cv_select_knn_k(splits$train, num_vars, cat_vars, k_grid, folds = 5, seed = 42)
cv_grids$knn <- knn_cv
best_k <- knn_cv$k[knn_cv$best][1]
best_hp$knn <- list(k = best_k)

val_probs$knn <- knn_probs(train_df, val_df, num_vars, cat_vars, best_k)
```

## Hyperparameter Tuning Plots

Parallel-coordinates plot of each tuned model's CV grid (`cv_grids`,
captured during fitting above) — one axis per hyperparameter plus
`mean_auc` — via `plot_cv_grid()` in `R/eval_plots.R`. Scales to any
number of hyperparameters (unlike a 2D heatmap), so it still works if
`rf`/`xgboost`/others grow beyond 2 tuned params.

```{r plot-cv-grids, fig.width=7, fig.height=5}
purrr::walk(names(cv_grids), function(name) {
  p <- plot_cv_grid(cv_grids[[name]], title = paste0("Hyperparameter Tuning — ", name))
  print(p)
})
```

# 7. Feature Importance (tree, rf, xgboost)

Only models with a native/built-in importance metric are shown here —
see `get_feature_importance()` in `R/importance.R`. `qda`/`knn` have no
single importance score, and the coefficient-based models (logistic,
logistic_complex, lda, glmnet) are already covered by `table_coef()` in
`R/tables.R`.

```{r plot-feature-importance, fig.width=7, fig.height=5}
importance_models <- c("tree", "rf", "xgboost")
purrr::walk(importance_models, function(name) {
  imp <- get_feature_importance(fitted_models[[name]])
  p <- plot_feature_importance(imp, title = paste0("Feature Importance — ", name))
  print(p)
})
```

# 8. Tune Threshold

## Ensemble (Mean Pool)

Simplest possible ensemble: mean-pool every model's predicted probability
per row (`ensemble_probs()` in `R/ensemble.R`), treat that pooled
probability like any other model's `probs` vector, and threshold-tune it
the same way (`optimal_threshold()`) — no separate meta-model, no weighting.
`ensemble_pruned` applies one basic pruning step first: rank models by
validation AUC (`prune_by_auc()`) and pool only the top 5, so a couple of
weak models can't drag down the pooled probability.

```{r ensemble}
base_val_probs <- val_probs
val_probs$ensemble <- ensemble_probs(base_val_probs)

pruned_names <- prune_by_auc(base_val_probs, actual_val, top_n = 5)
pruned_names
val_probs$ensemble_pruned <- ensemble_probs(base_val_probs[pruned_names])
```

## Validation Performance (all models, default 0.5 threshold)

Sanity check on the CV-selected final models before threshold tuning
(hyperparameters were already chosen via CV on train above, so this is no
longer part of model selection).

```{r val-comparison}
val_comparison <- compare_models(val_probs, actual_val)
table_model_comparison(val_comparison)
```

## Decision-Threshold Tuning (Validation Set)

Per-model F1-optimal threshold, chosen on validation only.

```{r threshold-tuning}
thresholds <- purrr::map_dbl(val_probs, ~ optimal_threshold(.x, actual_val, metric = "f1"))
thresholds
```

For each model, sweep the decision threshold from 0 to 1 and plot the
selected `metrics` (default sensitivity/specificity) against threshold,
marking the threshold that maximizes `target` ("f1" or "youdens_j",
default "youdens_j") with a vertical line and annotation. See
`plot_threshold_tuning()` in `R/eval_plots.R`.

```{r plot-threshold-tuning, fig.width=8, fig.height=5}
purrr::walk(names(val_probs), function(name) {
  p <- plot_threshold_tuning(val_probs[[name]], actual_val) +
    ggplot2::labs(title = paste0("Decision-Threshold Tuning — ", name))
  print(p)
})
```

# 9. ROC Curves — All Models

```{r test-probs}
test_probs <- purrr::imap(fitted_models, function(model, name) predict_fn(model, test_df))
test_probs$knn <- knn_probs(train_df, test_df, num_vars, cat_vars, best_k)

base_test_probs <- test_probs
test_probs$ensemble <- ensemble_probs(base_test_probs)
test_probs$ensemble_pruned <- ensemble_probs(base_test_probs[pruned_names])

actual_test <- as.integer(test_df$Status == "Fail")
```

```{r roc-all, fig.height=7}
plot_roc(test_probs, actual_test)
```

# 10. Final Evaluation

## Evaluate All Models on Test (Optimal Thresholds)

```{r test-comparison}
test_comparison <- compare_models(test_probs, actual_test, thresholds = thresholds)
table_model_comparison(test_comparison)
```

## Confusion Matrix — Best Model

```{r best-model-cm}
best_name <- test_comparison$model[which.max(test_comparison$f1)]
best_threshold <- thresholds[[best_name]]

best_eval <- evaluate_predictions(test_probs[[best_name]], actual_test, threshold = best_threshold)
plot_confusion_matrix(best_eval$confusion_matrix,
                       title = paste0("Confusion Matrix — ", best_name,
                                       " (threshold = ", round(best_threshold, 2), ")"))
```
