---
title: "Project 2 Applied Stats"
author: "Avery John"
date: "2026-07-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/averyjohn/Documents/Project2Stats", error = TRUE)
```

### Project Description / Notes
At least 5 Models need to be built for this project\
1) Additive Logistic Regression Model (Obj 1)\
2) Complex Logistic Regression Model (interactions or polynomials)\
3) LDA Model\
4) QDA Model\
5) Non-parametric Model (KNN, Decision Tree, Random Forest, etc.)\
\
Objective 1 Goals: Perform EDA, Build logistic regression model (additive is preferable) for interpretative purposes using training data\
Objective 2 Goals: Use simple logistic regression model as the baseline to build additional and more complex models to improve on prediction performance metrics\
\
OVERALL GOAL: Choose the model with the best predictive performance\
- AIC, ROC curves, LogLoss\

### Load Data
```{r}
student <- read.csv("StudentPerformanceFactors.csv", header = TRUE)
head(student)
```

### Load Libraries
```{r}
library(naniar)
library(tidyverse)
library(ggplot2)
library(GGally)
library(dplyr)
```

### Step 1 - Sanity Check (Assess the data)
```{r}
nrow(student)
str(student)
colSums(is.na(student))
```

```{r}
# I didn't notice any NA values until I looked at the levels for Distance_From_Home
table(student$Distance_from_Home)

# 67 values don't have a label, effectively making them NA values
# Perform a sanity check on other variables to check for the same issue
table(student$Parental_Involvement) #Good
table(student$Access_to_Resources) #Good
table(student$Extracurricular_Activities) #Good
table(student$Motivation_Level) #Good
table(student$Internet_Access) #Good
table(student$Family_Income) #Good
table(student$Teacher_Quality) # "" Column (78)
table(student$School_Type) #Good
table(student$Peer_Influence) #Good
table(student$Learning_Disabilities) #Good
table(student$Parental_Education_Level) # "" Column (90)
table(student$Gender) #Good
table(student$Physical_Activity)

# Drop NAs ("" values)
student_clean <- student %>%
  mutate(across(c(Teacher_Quality, Parental_Education_Level, Distance_from_Home), ~ na_if(., ""))) %>%
  drop_na(Teacher_Quality, Parental_Education_Level, Distance_from_Home)

nrow(student_clean)  # Dropped 229 rows, even though there are 235 rows with "" (67+78+90)

```

### Step 2: Convert categorical variables into factors
```{r}
student_clean <- student_clean %>%
  mutate(
    # Ordinal (Low < Medium < High, etc.)
    Parental_Involvement = factor(Parental_Involvement, levels = c("Low","Medium","High"), ordered = TRUE),
    Motivation_Level = factor(Motivation_Level, levels = c("Low","Medium","High"), ordered = TRUE),
    Access_to_Resources = factor(Access_to_Resources, levels = c("Low","Medium","High"), ordered = TRUE),
    Family_Income = factor(Family_Income, levels = c("Low","Medium","High"), ordered = TRUE),
    Teacher_Quality = factor(Teacher_Quality, levels = c("Low","Medium","High"), ordered = TRUE),
    Parental_Education_Level = factor(Parental_Education_Level, levels = c("High School","College","Postgraduate"), ordered = TRUE),

    # Nominal (no inherent order)
    Extracurricular_Activities  = factor(Extracurricular_Activities),
    Internet_Access = factor(Internet_Access),
    School_Type = factor(School_Type),
    Peer_Influence = factor(Peer_Influence),
    Learning_Disabilities = factor(Learning_Disabilities),
    Distance_from_Home = factor(Distance_from_Home),
    Gender = factor(Gender)
  )

str(student_clean)
```

### Step 3: Create binary response variable
```{r}
student_clean <- student_clean %>%
  mutate(Pass = factor(if_else(Exam_Score >= 67, "Pass", "Fail"), levels = c("Fail","Pass")))

table(student_clean$Pass)
prop.table(table(student_clean$Pass))

# 67 was chosen as the grade threshold, because if I used a higher threshold, more students would fail rather than pass
# Example:
#student_clean <- student_clean %>%
  #mutate(Pass = factor(if_else(Exam_Score >= 70, "Pass", "Fail"), levels = c("Fail","Pass")))

#table(student_clean$Pass)
#prop.table(table(student_clean$Pass)) 
    #75% fail while 25% pass

str(student_clean)
```

### Quick Outlier Check
```{r}
summary(student_clean$Exam_Score) #55 is the lowest score, while 101 is the highest (either data entry error or bonus points added)

boxplot(student_clean$Exam_Score, main = "Exam Score Distribution")
student_clean %>% filter(Exam_Score > 95)
# Dropping the 101 grade to stay on the average 100 point scale
student_clean <- student_clean %>%
  filter(Exam_Score <= 100)

nrow(student_clean)  # Now 6377 observations
summary(student_clean$Exam_Score) # Max is now 100.00
```

### Step 4: Univariate EDA
```{r}
# Numeric Variables
num_vars <- c("Hours_Studied", "Attendance", "Sleep_Hours", 
              "Previous_Scores", "Tutoring_Sessions",
              "Physical_Activity","Exam_Score")

student_clean %>%
  select(all_of(num_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Numeric Variables")

student_clean %>%
  select(all_of(num_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "lightblue") +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Boxplots of Numeric Variables")
```

```{r}
# Categorical Variables
cat_vars <- c("Parental_Involvement", "Access_to_Resources",
              "Extracurricular_Activities", "Motivation_Level", 
              "Internet_Access", "Family_Income", "Teacher_Quality",
              "School_Type", "Peer_Influence", "Learning_Disabilities", 
              "Parental_Education_Level", "Distance_from_Home", "Gender")

student_clean %>%
  select(all_of(cat_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value",
               values_transform = as.character) %>%
  ggplot(aes(x = value)) +
  geom_bar(fill = "coral") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Categorical Variables")
```

```{r}
# Response Variable (Pass)
student_clean %>%
  ggplot(aes(x = Pass, fill = Pass)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Distribution of Pass/Fail", y = "Count") +
  theme(legend.position = "none")
```

### Step 5: Bivariate EDA
```{r}
# Numeric Variables
student_clean %>%
  select(all_of(num_vars), Pass) %>%
  select(-Exam_Score) %>%  # exclude Exam_Score since Pass is derived from it directly
  pivot_longer(-Pass, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = Pass, y = value, fill = Pass)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Numeric Predictors by Pass/Fail")
```

```{r}
student_clean %>%
  select(all_of(num_vars), Pass) %>%
  select(-Exam_Score) %>%
  group_by(Pass) %>%
  summarise(across(everything(), mean, na.rm = TRUE))

# Attendance and Hours_Studied seem to be significant and related to attendance
# Physical Activity, Sleep_Hours and Tutoring_Sessions seem less significant
```

```{r}
# Categorical Variables
student_clean %>%
  select(all_of(cat_vars), Pass) %>%
  mutate(across(all_of(cat_vars), as.character)) %>%
  pivot_longer(-Pass, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = Pass)) +
  geom_bar(position = "fill") +
  facet_wrap(~ variable, scales = "free_x", ncol = 3) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Pass Rate by Categorical Variable", y = "Proportion")
```

```{r}
# This plot is to test a potential interaction between Hours Studied and Parental Involvement, seeing how the pattern spans across all levels
student_clean %>%
  ggplot(aes(x = Hours_Studied, fill = Pass)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Parental_Involvement) +
  theme_minimal() +
  labs(title = "Hours Studied Distribution by Pass/Fail, across Parental Involvement")

student_clean %>%
  group_by(Parental_Involvement) %>%
  summarise(
    mean_fail = mean(Hours_Studied[Pass == "Fail"]),
    mean_pass = mean(Hours_Studied[Pass == "Pass"]),
    gap = mean_pass - mean_fail,
    sd_pooled = sd(Hours_Studied)
  )

# Low seems to behave different from Medium and High, who behave similarly. 
# Plausible pattern worth testing as an interaction in objective 2.
```

```{r}
# Running the same standardized gap check for categorical variables

# Function: for a given categorical variable, compute the standardized Hours_Studied gap between Pass and Fail within each level
gap_check <- function(catvar) {
  student_clean %>%
    group_by(.data[[catvar]]) %>%
    summarise(
      mean_fail  = mean(Hours_Studied[Pass == "Fail"]),
      mean_pass  = mean(Hours_Studied[Pass == "Pass"]),
      gap        = mean_pass - mean_fail,
      sd_pooled  = sd(Hours_Studied),
      std_gap    = round(gap / sd_pooled, 3),
      n          = n(),
      .groups = "drop"
    ) %>%
    rename(level = 1) %>%
    mutate(level = as.character(level)) %>%
    mutate(variable = catvar, .before = 1)
}

gap_results <- map_dfr(cat_vars, gap_check)
gap_results %>% 
  select(variable, level, n, mean_fail, mean_pass, gap, std_gap) %>%
  knitr::kable(digits = 3)

gap_range <- gap_results %>%
  group_by(variable) %>%
  summarise(
    min_std_gap = min(std_gap),
    max_std_gap = max(std_gap),
    range_std_gap = round(max_std_gap - min_std_gap, 3),
    .groups = "drop"
  ) %>%
  arrange(desc(range_std_gap))

gap_range
```

### STEP 5 TAKEAWAYS:
`Parental_Involvement` and `Motivation_Level` stand out as the two strongest candidates for interaction terms. There is a clear separation in range, and both have reasonably large groups at every level.\
\
`Learning_Disabilities` looks promising (ranked at 3rd) but proceed with caution, as the "Yes" group (n = 668) is noticeably smaller than the "No" group (n = 5,709). That gap could partly reflect sampling noise rather than a true interaction. Still worth testing formally, just don't over-weight it from this screening alone.\
\
`Gender` and `Extracurricular_Activities` show almost no variation in `std_gap` across their levels (0.012 and 0.007). There is virtually no evidence of an interaction with `Hours_Studied` here. These variables are safe to leave as simple additive terms (no interaction).\
\
Everything else in the middle (`Teacher_Quality`, `Parental_Education_Level`, `Peer_Influence`, etc.) shows a modest but real signal. Not as significant as the top 2.

### Step 6: Multicollinearity Check
```{r}
# Numeric Predictors
student_clean %>%
  select(all_of(num_vars)) %>%
  cor() %>%
  round(3)

# Noticeable: Moderately high correlation between Attendance and Exam_Score (0.582) and Hours_Studied and Exam_Score (0.446). Exam_Score is the common denominator.
```

```{r}
student_clean %>%
  select(all_of(num_vars), -Exam_Score) %>%
  ggpairs(
    lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5)),
    diag = list(continuous = wrap("densityDiag")),
    upper = list(continuous = wrap("cor", size = 4))
  ) +
  theme_minimal()

# There is no pair of predictors that are correlated with one another according to this plot. All values are close to 0.
```

```{r}
# Categorical Predictors - Chi-sq Independence Test
table(student_clean$Family_Income, student_clean$Access_to_Resources) %>%
  prop.table(margin = 1) %>%
  round(3)

chisq.test(table(student_clean$Family_Income, student_clean$Access_to_Resources))

# Test Results: There's no statistical evidence of an association between Family_Income and Access_to_Resources in this data. (p-value = 0.4071)
# We failed to reject the null hypothesis of independence between the two variables 
# The observed proportions are nearly identical across income levels, suggesting no meaningful multicollinearity concern between these two variables.
```

```{r}
# Creating a loop to run a chi-square test across all pairwise combinations of categorical variables

cat_pairs <- combn(cat_vars, 2, simplify = FALSE)

chisq_results <- map_dfr(cat_pairs, function(pair) {
  var1 <- pair[1]
  var2 <- pair[2]
  
  tbl <- table(student_clean[[var1]], student_clean[[var2]])
  
  test <- suppressWarnings(chisq.test(tbl))
  
  tibble(
    var1     = var1,
    var2     = var2,
    chi_sq   = round(unname(test$statistic), 3),
    df       = unname(test$parameter),
    p_value  = round(test$p.value, 4),
    cramers_v = round(sqrt(unname(test$statistic) / (sum(tbl) * (min(dim(tbl)) - 1))), 3)
  )
})

chisq_results %>% arrange(desc(cramers_v)) 

#Cramer's V = statistical measure of the strength of association between two categorical variables, ranging from 0 (no association) to 1 (perfect association) -- https://www.statology.org/interpret-cramers-v/

# Cramer's V Thresholds based upon DF:
# df*	Small	Medium	Large
# 1	  0.10	0.30	  0.50
# 2	  0.07	0.21	  0.35
# 3	  0.06	0.17	  0.29
# 4	  0.05	0.15	  0.25

# 0.2 sits in the middle of the Medium range, so that's the value chosen as the threshold

chisq_results %>%
  filter(cramers_v > 0.2) %>%
  arrange(desc(cramers_v))
```

### STEP 6 TAKEAWAYS:
Running a chi-square test of independence including Cramer's V across all 78 categorical variable combinations resulted in a maximum association of V = 0.034 (Access_to_Resources vs. School_Type). This indicates negligible multicollinearity risk among categorical predictors, so multicollinearity is not a concern with all the categorical variable pairs.\ 
\
This is corroborated by the earlier chi-square test between Family_Income and Access_to_Resources (p = 0.407), which similarly showed no meaningful relationship.\
\
Even if the smallest threshold (0.05) was chosen, our results will still be the same, as 0.034 is the max Cramer's V on the list. This supports our conclusion of no multicollinearity between catagorical variables even further.


